---
title: "Final Project Activity 1"
author: "Parker Kuchulan & Bradley Monarch"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, eval = TRUE,
                      message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(lubridate)
library(skimr)
```


## Import Data

```{r}
gunData <- data.table::fread("gunViolenceFull.csv")
censusData <- data.table::fread('https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/census_data_2008-2021.csv')
```

## Explore and display high-level characteristics of your data set, e.g., important variables and their types, different levels for factor variables, any patterns of missing values.

```{r}
summary(gunData)
glimpse(gunData)
skim(gunData)
```

## Obtain at least two tables of group-level summary statistics for quantitative variables in your data set.

```{r}
#Summary statistics for number of people killed
gunData %>% pull(n_killed) %>% summary()
```

```{r}
#Summary statistics for number of people injured
gunData %>% pull(n_injured) %>% summary()
```

## Obtain at least one frequency table showing the cross-section of two categorical variables in your data set.
```{r}
#Freq table for state totals of killed and injured
stateCount <- gunData %>% mutate(fatal = n_killed > 0) %>% 
  mutate(type = ifelse(fatal == FALSE, "Injured", "Killed"))
stateCount %>% count(state, fatal)
```

```{r}
#New dataset for properly formated dates
gunDataT <- stateCount %>% mutate(year = lubridate::year(date)) 

#Remove years with incomplete data
gunDataF <- gunDataT %>% filter(year != 2013 & year != 2018)

#Remove years from censusData to match gunData
censusDataF <- censusData %>% filter(year > 2013 && year < 2018)
#censusData %>% pull(year) %>% skim()
#censusData %>% count(year)
```

## Create one ggplot visualization for your data, including appropriate labels with units of measurement when applicable and a specified theme.

```{r}
#Number of people killed in shootings by year
gunDataF %>% ggplot(aes(x = year,
                        y = sum(fatal))) + 
  geom_col() +
  labs(title = "Number of people killed in shootings by year",
       x = "Year",
       y = "Total people killed") + 
  theme_bw() +
  scale_y_continuous(labels = scales::comma)
```

## New dataframe with state death totals
```{r}
stateVictims <- gunData %>%
  group_by(state) %>% 
  summarize(totalKilled = sum(n_killed), totalInjured = sum(n_injured))
```


## Creating a GGplot visualization of a bar chart of the states with the top 10 states in number of victims killed 
```{r}
stateVictims %>% arrange(desc(totalKilled)) %>%
  slice_head(n = 10) %>% 
  ggplot(aes(x = reorder(state, +totalKilled), y = totalKilled)) +
  geom_col(fill = "#800020", color = "black") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 8), 
        legend.position = "none") +
  scale_y_continuous(breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000), limits = c(0, 6000)) +
  labs(title = "Top Ten States with the Most Fatalities from Gun Violence (2013-2018)", x = "State", y = "Total Fatalities", caption = "Data Source: Gun Violence Archive (GVA)")
```

## Creating Data Dictionary
```{r}
# Creating data dictionary. 
dataDictionary <- tibble(Variable = colnames(gunDataT),
                         Description = c("gunviolencearchive.org ID for incident",
                                         "date of occurrence",
                                         "state",
                                         "city or county",
                                         "address where incident took place",
                                         "number of people killed",
                                         "number of people injured",
                                         "link to gunviolencearchive.org webpage containing details of incident",
                                         "link to online news story concerning incident",
                                         "ignore, always False",
                                         "Congressional district",
                                         "gun stolen or not, e.g. 'Unknown' or 'Stolen'",
                                         "description of gun type",
                                         "list of incident characteristics",
                                         "latitude of location",
                                         "description of location where incident took place",
                                         "longitude of location",
                                         "number of guns involved",
                                         "additional notes about the incident",
                                         "participant age",
                                         "participant description of age group, e.g. 'Adult 18+'",
                                         "participant gender being 'Male' or 'Female'",
                                         "participant name",
                                         "relationship of participant to other participants",
                                         "outcome, one of 'Arrested', 'Killed', 'Injured', or 'Unharmed'",
                                         "participant category being 'Victim' or 'Subject-Suspect'",
                                         "links to online news stories concerning incident",
                                         "state house district",
                                         "state senate district",
                                         "full address", "whether the victim was killed or not", "whether the victim was killed or not as a string", "the year of the incident"),
                         Type = map_chr(gunDataT, .f = function(x){typeof(x)[1]}),
                         Class = map_chr(gunDataT, .f = function(x){class(x)[1]}))
```


```{r}
#Initialize dictionary
flextable::flextable(dataDictionary, cwidth = 2)
```
